---
title: "Age of the Congress representative"
author: "Huiting Wu"
date: "2023-12-04"
output:
  html_document:
    self_contained: true
---

```{r, message=FALSE}
library(ggplot2)
library(readr)
library(readxl)
library(dplyr)
library(maps)
library(mapproj)
library(lubridate)
library(gridExtra)
library(tidyverse)
library(gganimate)
library(gifski)
library(transformr)
```

## Data Resources

1. **US percentage population of age groups over time dataset:**

- Description: Persons in selected age groups as a percentage of the total U.S. population and children ages 0–17 as a percentage of the dependent population, 1950–2022 and projected 2023–2050.

- Website: github

- Link: <https://www.childstats.gov/americaschildren/tables/pop2.asp>

2. **Aging congress dataset:**

- Description: This directory contains various demographic data about the United States Senate and House of Representatives over time. It’s been used in the following FiveThirtyEight articles: Congress Today Is Older Than It’s Ever Been, by Geoffrey Skelley (April 3, 2023).
The dataet contains information about the age of every member of the U.S. Senate and House from the 66th Congress (1919-1921) to the 118th Congress (2023-2025). Data is as of March 29, 2023, and is based on all voting members who served in either the Senate or House in each Congress. The data excludes delegates or resident commissioners from non-states. Any member who served in both chambers in the same Congress was assigned to the chamber in which they cast more votes. We began with the 66th Congress because it was the first Congress in which all senators had been directly elected, rather than elected by state legislatures, following the ratification of the 17th Amendment in 1913.

- Website: Childstats

- Link: <https://github.com/fivethirtyeight/data/blob/master/congress-demographics/data_aging_congress.csv>

3. **State Label**

- Description: It contains name of USA states and its abbreviation.

- website: github

- Link: https://github.com/jasonong/List-of-US-States/blob/master/states.csv

## Data Import and Data Information

```{r, message=F}
pop <- read_excel("pop2.xlsx")

create_and_transpose_table <- function(row, col, dataset) {
  table <- dataset[row, col]
  table <- as.data.frame(t(table))
  colnames(table) <- as.character(table[1, ])
  table <- table[-1, ]
  return(table)
} # create a function to organize the tables

table1 <- create_and_transpose_table(1:7, 1:21, pop)
table2 <- create_and_transpose_table(9:15, 1:21, pop)
table3 <- create_and_transpose_table(17:23, 1:21, pop)
table4 <- create_and_transpose_table(25:31, 1:21, pop)
pop_table <- bind_rows(table1, table2, table3, table4)

pop_table <- within(pop_table, 
            rm("Percentage of total population", 
            "Children ages 0–17 as a percentage of the dependent populationa"))
pop_table <- mutate_all(pop_table, as.numeric)
pop_table <- mutate_all(pop_table, as.integer)
pop_table <- pop_table[-74:-80, ] 
# remove the prejected years from the dataset

head(pop_table)
```

```{r, message=FALSE}
congress <- read_csv("data_aging_congress.csv")

congress <- congress |> mutate(party = case_when(
    party_code == 200 ~ 'Republicans',
    party_code == 100 ~ 'Democrats',
    TRUE ~ as.character(NA)
  ))
# there are more party code, but I only focus on Republicans and Democrats

congress <- congress |> mutate(age_group = case_when(
  generation == "Gilded" ~ "1822:1842",
  generation == "Progressive" ~ "1843:1859",
  generation == "Missionary" ~ "1860:1882",
  generation == "Lost" ~ "1883:1900",
  generation == "Greatest" ~ "1901:1927",
  generation == "Silent" ~ "1928:1945",
  generation == "Boomers" ~ "1946:1964",
  generation == "Gen X" ~ "1965:1980",
  generation == "Millennial" ~ "1981:1996",
  generation == "Gen Z" ~ "1997:2012",
  TRUE ~ as.character(NA)
))

head(congress)
```

```{r, message=FALSE}
state_names <- read_csv("states.csv")
head(state_names)
```


```{r}
dim(pop_table)
```

There are totel 105 rows and 5 columns in the US percentage population of age groups over time dataset. After remove the predicted age from the dataset, there are 73 rows on the pop_table.

```{r}
dim(congress)
```
There are totle 29120 rows and 13 columns in the aging congress dataset. I created 2 new columns into the original dataset which are age_group and party. So there are 15 columns.

```{r}
dim(state_names)
```
There are totel 51 rows and 2 columns in the state names data set. This dataset includes D.C. as a state.

## Variable Description of datasets

1. **US percentage population of age groups over time dataset:**

* congress: The number of the Congress that this member’s row refers to.

* start_date: First day of a Congress.

* chamber: The chamber a member of Congress sat in: Senate or House. 

* state_abbrev: The two-letter postal abbreviation for the state a member represented.

* party_code: A code that indicates a member’s party, based on the system used by the Inter-university Consortium for Political and Social Research. The most common values will be 100 for Democrats, 200 for Republicans.

* bioname: Full name of member of Congress.

* bioguide_id: Code used by the Biographical Directory of the United States Congress to uniquely identify each member.

* birthday: Date of birth for a member.

* cmltv_cong: The cumulative number of Congresses a member has or had served in (inclusive of listed congress), regardless of whether the member was in the Senate or House. 

* cmltv_chamber: The cumulative number of Congresses a member has or had served in a chamber (inclusive of listed congress). 

* age_days: Age in days, calculated as start_date minus birthday.

* age_years: Age in years, calculated by dividing age_days by 365.25.

* generation: Generation the member belonged to, based on the year of birth. 

New variables I added:

* age_group: Convert the Generations in the data to the defined year range

* party: Convert the party code to the name of the party which the code represent.

2. **Aging congress dataset:**

Age group as a percentage of total population:

* ages 0-17

* ages 18-64

* ages 65 and older

Age group as a percentage of the dependent population:

* ages 0-17

3. **State Label**

* State

* Abbreviation


## Research Questions

* What is the trend of the age of the Congress representative change over time?

* Does the changing of age of Congress representative related with the age of population?

* Which state with the oldest Congress representative on average?

## Visualization

### Using Congress age dataset: Create an animation time series plot to show the trend of the age of the Congress representative change over time

```{r}
congress |> 
  group_by(Year = year(start_date)) |> 
  summarise(count = n(),
  Age = mean(age_years)) |>
  filter(Age == max(Age)| Age == min(Age))
```

```{r, warning=FALSE}
class(congress$start_date)
graph_data <- congress |> 
  group_by(Year = year(start_date)) |> 
  summarise(count = n(),
  Age = mean(age_years, na.rm = T))

graph_data$Year <- as.numeric(graph_data$Year)

anim <- ggplot(graph_data, aes(x = Year, y = Age, color = Age)) +
  geom_point() +
  transition_time(Year) +
  shadow_wake(wake_length = 5, alpha = FALSE) +
  scale_color_gradient(name = "Average Age", 
                       low = "green", high = "red") +
  geom_hline(yintercept = mean(congress$age_years, na.rm = T), 
             linetype = "dashed", 
             color = "gray") +
  annotate("text", x = max(congress$year), 
           y = mean(congress$age_years, na.rm = T),
           label = sprintf("Overall Average Age: %.2f", mean(congress$age_years, na.rm = T)),
           hjust = -1.13, vjust = -1.5, 
           color = "orange", size = 3.75) +
  annotate("text", x = 1981,
           y = 49.52,
           label = sprintf("Lowest Avg Age: %.2f", 49.51690),
           color = "green", size = 3.75) + 
  annotate("text", x = 2021,
           y = 58.83,
           label = sprintf("Highest Avg Age: %.2f", 58.83102),
           hjust = 1, vjust = -0.5,
           color = "red", size = 3.75) +
  labs(y = "Average age",
       x = "Year",
       title = "What is the trend of the age of the Congress change over time?") +
  theme_classic()
anim
```
The average age of the congress representative is about 53.73 years old. Before the year 1981, the average age of the congress representative is around the overall average age. Around the year 1975, there is a big decreasing. In the year 1981, the average age of congress representative is the lowest which is about 49.52 yeas old. After the year 1981, the average age of the congress representative is increasing above the overall average age about 5 years.

### Using proportion of US population for each age groups over time dataset: Create a time serise plot to show the change of proportion of population age above 18 in the USA over time.

```{r, warning=FALSE}
pop_table$pop_18 <- pop_table[,3] + pop_table[,4]
ggplot() + 
  geom_line(data = pop_table, 
            aes(x = Age, y = pop_18, color = pop_18)) +
  scale_color_gradient(name = "Proportion of population over 18 years old", 
                       low = "green", high = "red") +
  geom_hline(yintercept = mean(pop_table$pop_18, na.rm = T), 
             linetype = "dashed", 
             color = "gray") +
  geom_text(aes(x = max(pop_table$Year, na.rm = T), 
                y = mean(pop_table$pop_18, na.rm = T),
                label = sprintf("Overall Average Proportion 
of population over 18 years old: %.2f", mean(pop_table$pop_18, na.rm = T))),
                hjust = -0.1, vjust = 0.5, 
                color = "orange", size = 3.75) +
  labs(x = "Year",
       y = "Proportion of population age above 18 in the USA",
       title = "Proportion of Adults in USA over Years") +
  theme_classic()
```

By the graph above, there is an increasing in proportion of adults over 18 years old in the USA after about the year of 1960. After the year about 1980, the proportion of adults over 18 years old is higher than the overall average proportion. Previously, the average age of the congress representative is increasing above the overall average age also in after the year 1981.

```{r, warning=FALSE}
year_age_graph <- congress |> 
  group_by(Year = year(start_date)) |> 
  summarise(count = n(),
  Age = mean(age_years)) |> 
  ggplot(aes(x = Year, y = Age)) +
  geom_line(aes(color = Age)) +
  scale_color_gradient(name = NA, 
                       low = "green", high = "red") +
  geom_hline(yintercept = mean(congress$age_years), 
             linetype = "dashed", 
             color = "gray") +
  geom_text(aes(x = max(congress$year), 
                y = mean(congress$age_years),
                label = sprintf("Overall Average Age: %.2f", mean(congress$age_years))),
                hjust = -1.13, vjust = -1.5, 
                color = "orange", size = 3.75) +
  labs(y = "Average age",
       x = "Year") +
  theme_classic() +
  guides(color = FALSE)

year_pop_graph <- ggplot() + 
  geom_line(data = pop_table, 
            aes(x = Age, y = pop_18, color = pop_18)) +
  scale_color_gradient(name = NA, 
                       low = "green", high = "red") +
  geom_hline(yintercept = mean(pop_table$pop_18, na.rm = T), 
             linetype = "dashed", 
             color = "gray") +
  geom_text(aes(x = max(pop_table$Year, na.rm = T), 
                y = mean(pop_table$pop_18, na.rm = T),
                label = sprintf("Overall Average Proportion 
of population over 18 years old: %.2f", mean(pop_table$pop_18, na.rm = T))),
                hjust = -0.1, vjust = 0.5, 
                color = "orange", size = 3.75) +
  labs(x = "Year",
       y = "Proportion of Adults",
       title = "Does the changing of age of Congress representative 
       related with the age of population?") +
  theme_classic() +
  guides(color = FALSE) +
  xlim(1919, 2023)
```


### Using proportion of US population for each age groups over time dataset and aging congress dataset: Combined the two plots to find out the relationship between age of population and age of congress representative
```{r, warning=FALSE}
grid.arrange(year_pop_graph, year_age_graph, ncol= 1)
```

The top graph, the y-axis represents the proportion of Adults in USA (age over 18). The bottom graph, the y-axis represents the average age of congress representative. The x-axis for both graphs is the year from 1919 to 2023. Both of the graphs has a similar trend but the deep is in a different year. By comparing of these two graphs, there is a similar trend of increase after the year 2000. In 2000, the proportion of adults in USA is increasing and the average age of the congress representative is increasing also.
The age of the population has some kind of relationship with the aging of congress representative after 2000. With a larger adults age group in the USA, the age of the congress representative is increasing.

### USA State Map Data
```{r}
states <- map_data("state")
states <- states |> 
  rename(State = region)
states <- states |> 
  select(-subregion)
```

### Using Congress age data and State label Data

```{r, warning=F, message=FALSE}
joined_data <- left_join(congress, state_names, 
                         by = c("state_abbrev" = "Abbreviation"))

joined_data <- joined_data |> 
  mutate(State = tolower(State)) |> 
  filter(!(state_abbrev %in% c("HI", "AK")))

congress_with_map_data <- left_join(states, joined_data, by = "State") |> 
  select(long, lat, age_years, State, state_abbrev) |> 
  group_by(State = as.factor(State), 
           state_abbrev = as.factor(state_abbrev)) |> 
  summarise(long = mean(long),
            lat = mean(lat),
            Avg_age = mean(age_years, na.rm = TRUE)) |> 
  filter(!is.na(Avg_age))
# There is a missing value for district of columbia. I choose to remove it.
```

#### Bar Chart: To show the average age of congress representative from each states. The height of the bar can easily oberved the difference.

```{r}
ggplot(congress_with_map_data, aes(x = reorder(State,-Avg_age), y = Avg_age)) + 
  geom_bar(stat = "identity", fill = "lightblue", color = "black") +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1)) +
  labs(x = "State", y = "Average age", 
       title = "Which state with the oldest Congress representation on average?")
```

From the bar chart, average age of the congress representative for the state Vermont is the older than others and Wyoming state is the second one. 

#### Main USA Map with Average Age of Congress Representative for each states: Using the map to show the location of the states and visual the age of Congress Representative for each states.

```{r}
state_labels <- congress_with_map_data %>%
   group_by(state_abbrev) %>%
   summarise(
      long = median(long), 
      lat = median(lat)
   )

ggplot() +
  geom_polygon(data = states, aes(x = long, y = lat, group = group),
  fill = "lightblue", color = "black") +
    geom_point(data = congress_with_map_data, 
               aes(x = long, y = lat, 
                   size = Avg_age, 
                   color = Avg_age)) +
  geom_text(data = state_labels, 
            aes(x = long, y = lat, 
                label = state_abbrev, 
                fontface = "bold"),
                size = 3, color = "white") +
  scale_color_gradient(name = "Average Age", 
                       low = "green", high = "red") +
  labs(title = "Average age of congress representative from each state",
       size = "Average Age",
       color = "Average Age") +
  coord_map() +
  theme_classic()
```

The dots with closer to red color and larger size represent older average age. We can see the dots for Vermont and Wyoming state are the larger and closer to red than the others state.


